%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development. Uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[disable,textsize=tiny]{todonotes}

% Paper-specific packages and macros (kept from prior preamble)
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{xspace}
\newcommand{\ie}{\emph{i.e.}\xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\SK}{\mathbf{S}_K}
\newcommand{\SR}{\mathbf{S}_R}
\newcommand{\dK}{\mathbf{d}_K}
\newcommand{\dR}{\mathbf{d}_R}
\newcommand{\hB}{\mathbf{h}_B}
%\newcommand{\todo}[1]{\textcolor{red}{[Note: #1]}}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{The Dark Subspace of Fine-Tuning Memorisation}

\begin{document}

\twocolumn[
  \icmltitle{The Dark Subspace of Fine-Tuning Memorisation}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package above.
  \icmlsetsymbol{equal}{*}

  % Anonymous double-blind submission. Replace with author block at camera-ready.
  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Submission}{anon}
  \end{icmlauthorlist}

  \icmlaffiliation{anon}{Anonymous Institution}

  \icmlcorrespondingauthor{Anonymous Submission}{anon@example.com}

  % You may provide any keywords that you find helpful for describing your
  % paper. These are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Membership Inference, Sparse Autoencoders, Fine-Tuning, Memorisation, Mechanistic Interpretability, Privacy Auditing}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...
\printAffiliationsAndNotice{}  % no special notice (required even if empty)

\begin{abstract}
Sparse autoencoders (SAEs) are increasingly used to interpret and edit language-model activations, including in privacy applications that assume the information leaked by training data sits inside the learned dictionary. We test this assumption for fine-tuned transformers. We split the model's hidden-state activations into the SAE reconstruction and the reconstruction residual, and we train membership detectors on each part to ask which part actually carries the leakage signal. In controlled fine-tuning experiments on Pythia, detectors recover more membership signal from the reconstruction residual than from the SAE reconstruction. An evaluation across nine models with an explicit reconstruction-quality validity gate shows the same residual-over-reconstruction ordering wherever the SAE is reliable enough to interpret, and the ordering is not explained by SAE initialization, detector fitting, dictionary-training data, or simple lexical artifacts. SAE feature interventions that operate only on the dictionary may therefore leave membership evidence detectable in the reconstruction residual.
\end{abstract}

\input{intro}
\input{related}
\input{methods}
\input{results}

\section{Concluding Remarks}
\label{sec:concluding}

\textbf{SAE feature usefulness and membership privacy are different properties.}
SAE reconstruction can weaken membership detection while the underlying evidence remains recoverable from the residual. Features remain useful sparse coordinates for interpreting or editing model outputs, but a reduction in measured detection does not mean the evidence has been removed. Any SAE feature-level intervention should therefore be paired with a residual audit before it supports a privacy claim.

\textbf{Scope and limitations.}
The controlled Pythia experiments carry the main causal evidence. The cross-architecture experiments support qualitative generality at the validity gate, and the norm-direction split and residual-PC ablation are descriptive, not causal. The audit is white-box and uses controlled OpenWebText fine-tuning with standard L1-regularized ReLU SAEs. Pre-training MIA, confounded by distribution shift~\citep{duan2024membership}, is not addressed here. Further limitations are in Appendix~\ref{app:limitations_detail}.

\bibliography{references}
\bibliographystyle{icml2026}

\appendix
\input{appendix}

\end{document}
