%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development. Uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[disable,textsize=tiny]{todonotes}

% Paper-specific packages and macros (kept from prior preamble)
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{xspace}
\newcommand{\ie}{\emph{i.e.}\xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\SK}{\mathbf{S}_K}
\newcommand{\SR}{\mathbf{S}_R}
\newcommand{\dK}{\mathbf{d}_K}
\newcommand{\dR}{\mathbf{d}_R}
\newcommand{\hB}{\mathbf{h}_B}
%\newcommand{\todo}[1]{\textcolor{red}{[Note: #1]}}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{The Dark Subspace of Fine-Tuning Memorisation}

\begin{document}

\twocolumn[
  \icmltitle{The Dark Subspace of Fine-Tuning Memorisation}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package above.
  \icmlsetsymbol{equal}{*}

  % Anonymous double-blind submission. Replace with author block at camera-ready.
  \begin{icmlauthorlist}
    \icmlauthor{Anonymous Submission}{anon}
  \end{icmlauthorlist}

  \icmlaffiliation{anon}{Anonymous Institution}

  \icmlcorrespondingauthor{Anonymous Submission}{anon@example.com}

  % You may provide any keywords that you find helpful for describing your
  % paper. These are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Membership Inference, Sparse Autoencoders, Fine-Tuning, Memorisation, Mechanistic Interpretability, Privacy Auditing}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...
\printAffiliationsAndNotice{}  % no special notice (required even if empty)

\begin{abstract}
Sparse autoencoders (SAEs) are increasingly used to interpret and edit language-model activations, including in privacy applications that assume the information leaked by training data sits inside the learned dictionary. We test this assumption for fine-tuned transformers. We split the model's hidden-state activations into the SAE reconstruction and the reconstruction residual, and we train membership detectors on each part to ask which part actually carries the leakage signal. In controlled fine-tuning experiments on Pythia, detectors recover more membership signal from the reconstruction residual than from the SAE reconstruction. A nine-model sweep with an explicit reconstruction-quality validity gate shows the same residual-over-reconstruction ordering wherever the SAE is reliable enough to interpret, and the ordering is not explained by SAE initialization, detector fitting, dictionary-training data, or simple lexical artifacts. SAE feature interventions that operate only on the dictionary may therefore leave membership evidence detectable in the reconstruction residual.
\end{abstract}

\input{intro}
\input{related}
\input{methods}
\input{results}

\section{Concluding Remarks}
\label{sec:concluding}

\textbf{SAE reconstruction omits signal recoverable from the residual.}
SAE reconstruction can weaken layer-local membership detection while leaving membership evidence recoverable from the reconstruction residual. This separates what feature-level interventions edit from what membership detectors can read, and it is the practical implication this paper supports.

\textbf{Feature-level interventions do not imply membership privacy.}
The intervention results sharpen this distinction. Selected SAE features can affect extractable recall while leaving membership detection unchanged at the AUROC level reported in the body. Explicitly training the SAE to reconstruct the known $\dK$ direction restores that direction in the reconstruction, but broader membership
detectors still recover signal from the residual. The claim is therefore not complete causal localization of a membership subspace. It is a recoverability
claim. Membership evidence remains decodable after the SAE decomposition, and residual evidence must be evaluated separately from feature-level output.

\textbf{Scope and limitations.}
The controlled Pythia experiments carry the primary causal evidence for the residual-over-reconstruction ordering. The cross-architecture sweep supports qualitative generality where the validity gate is satisfied, while the norm-direction split and residual-PC ablation are descriptive, not causal. The audit is white-box and uses controlled OpenWebText fine-tuning with standard L1-regularized ReLU SAEs. Pre-training MIA, confounded by distribution shift~\citep{maini2024dataset, duan2024membership}, is not addressed here. Held-out partition sensitivity, newer SAE variants, and other boundary conditions are detailed in Appendix~\ref{app:limitations_detail}.

\bibliography{references}
\bibliographystyle{icml2026}

\appendix
\input{appendix}

\end{document}
